A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-01 paper.tex
Some updates

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG, but at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the persistance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
\footnote{We use the standard graph theoretic sense of
an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but
where each individual is represented by two nodes (for diploids).
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

\section*{Event ARGs}\label{eARG}
We are mainly interested in exploring the properties of gARGs in
this manuscript, but in order to contrast with the existing
methods we concretely define an ``event ARG'', or eARG.

@@

In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

[TODO list the salient points that we want to contrast against
the gARG. We don't want this to seem like a straw man]

>>>>

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous [FIXME: this isn't really true,
what about parsimony ARGs?]
with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG, but at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] which persist over 
many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each 
edge\footnote{We use the standard graph theoretic sense of
an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
 is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but
where each individual is represented by two nodes (for diploids).
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
The final vital
detail is that we associate a \emph{breakpoint} with each recombination
event, and this information is sufficient to uniquely
define the local genealogical trees at every position along the genome.

As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).

% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.

% [TODO list the salient points that we want to contrast against
% the gARG. We don't want this to seem like a straw man]

% Griffiths and colleagues~\citep{griffiths1997ancestral} originally defined 
% the ARG as analog of the Kingman coalescent~\citep{kingman1982coalescent},
% in which the ancestry of a sample of $n$ genomes under idealised 
% conditions is modelled. One of the key assumptions of this model is that 
% the sample size $n$ is much less than the population size $N$, which means 
% that the probability of two multiple concurrent events (i.e., coalescence
% and recombination events) is negligible.

@@

[TODO: what do we need to cover here?]

[What is the ARG definition provided by the parsimony ARG literature?]

[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
quirky dependence on SNP data?]
